\thetitle

### 2021-03-03 writing/challenges.tex
dangling lines

Besides the run-time checks, the usefulness of blame might be due to pure luck.
Indeed, a proper evaluation of blame needs to consider the role of chance in blame assignment, too.
A random rational programmer mode provides yet another necessary baseline using the same ingredients as for exceptions.

>>>>

An experiment must rule out that the usefulness of blame assignment is sheer
luck.  Hence, a random rational programmer mode provides yet another necessary
baseline using the same ingredients as for exceptions.

### 2021-03-03 writing/mutate-benchmarks.tex
touch ups in section 6
dangling lines

The bar for testing the rational programmer is high: it demands a
collection of realistic programs that
\begin{itemize}

@@

   \item  fail and contain type-level mistakes. 
\end{itemize}

No such benchmark suite exists.  Worse, no benchmark suite of
gradually typed programs with type-level mistakes exists at
all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
Python programs to evaluate their technique of fixing mistakes in type
annotations. Their collection does not come with type-level mistakes
in the code itself.}

However, over the past decade,  \citet{gtnffvf-jfp-2019} have curated a
collection of Typed Racket programs for systematically measuring the
implementation's performance. The benchmark suite consists of fully typed,
correct programs, written by a number of different authors, who have
maintained and evolved these programs over several years. The programs
range widely in size, complexity, purpose, origin and in programming
style. The latter is mostly a reference to their reliance on many Typed
Racket features: occurrence typing~\cite{tf-icfp-2010}, types for mutable
and immutable data structures~\cite{hpst-sfp-2010}, types for first-class
classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
numeric tower~\cite{stathff-padl-12}.

>>>>

The benchmark programs for a rational-programmer experiment must 
\begin{enumerate}

@@

\item  fail and contain type-level mistakes. 
\end{enumerate}
\noindent
No such benchmark suite exists.  Worse, no benchmark suite of
gradually typed programs with type-level mistakes exists at
all.\footnote{\citet{cc-oopsla-20} created a collection of Reticulated
Python programs to evaluate their technique of fixing mistakes in type
annotations. Their collection does not come with type-level mistakes
in the code itself.}

Fortunately, \citet{gtnffvf-jfp-2019} have curated a collection of Typed Racket
programs for systematically measuring the implementation's performance. The
benchmark suite consists of fully typed, correct programs, written by a number
of different authors, who have maintained and evolved these programs over
several years. The programs range widely in size, complexity, purpose, origin
and in programming style. The latter is mostly a reference to their reliance on
many Typed Racket features: occurrence typing~\cite{tf-icfp-2010}, types for
mutable and immutable data structures~\cite{hpst-sfp-2010}, types for
first-class classes and objects~\cite{tsdtf-oopsla-2012}, and types for Racket's
numeric tower~\cite{stathff-padl-12}.

### 2021-03-03 writing/mutate-interesting.tex
touch ups in section 6
dangling lines

\item By definition, an impedance mismatch is a clash between the type ascription of one module's imports and another module's exports. Hence, type checking should fail for an interesting mutant.

@@

\item If the evaluation of a mutant module immediately raises an exception because
of the changes, there is no work for the rational programmer. Indeed, if the
stack trace contains source pointers to two modules, the scenario is still
uninteresting. Every ordinary benchmark program comes with a {\tt main} module
that acts as a driver, whose source is guaranteed to be included in the stack
trace.  Hence, the definition of interesting mutation insists on the presence of
three different modules in the stack trace. This guarantees that the debugging
scenario demands a sufficiently sophisticated effort, due to the interaction
between the buggy module with its context.  In these cases, the rational
programmer must contend with at least two modules involved in a faulty
interaction.

>>>>

\item An impedance mismatch is a clash between the type ascription of one
module's imports and another module's exports. Hence, type checking should fail
for an interesting mutant.

@@

\item If the evaluation of a mutated module immediately raises an exception because
of the changes, there is no work for the rational programmer. Indeed, if the
stack trace contains source pointers to two modules, the scenario is still
uninteresting. Every ordinary benchmark program comes with a {\tt main} module
that acts as a driver, whose source is guaranteed to be included in the stack
trace.  Hence, the definition of interesting mutation insists on the presence of
three different modules in the stack trace. This guarantees that the debugging
scenario demands a sufficiently sophisticated effort, due to the interaction
between the buggy module with its context.  In these cases, the rational
programmer must contend with at least two modules involved in a faulty
interaction.

### 2021-03-03 writing/results.tex
dangling lines

Figure~\ref{fig:success-bars} summarizes the overall success rates of
every mode.\footnote{The random mode behaves the same for
all three semantics, so this analysis considers it a single mode.}
The success rates illustrate a few points that underlie the
rest of the analysis.  The first notable piece of information from this
figure is that every mode has failed debugging scenarios, not just
Erasure. This should not come as a surprise to the astute reader.  Running
a rational programmer mode on a scenario may result in an exception
that carries no useful information about which module the programmer
should type next. For instance the stack trace of the exception may not
contain frames from any untyped module of the program. This can happen at any point
along a blame trail, causing it to fail.

@@

Figure~\ref{fig:avo-bars} also provides answers for $Q_*$.
Blame for all three semantics is significantly
more useful than Erasure exceptions (by around 35\%). Natural blame is
more useful than both versions of Transient blame by a small percentage;
in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
scenarios Transient blame is more useful. The Transient first and
Transient last blame are practically indistinguishable. Finally, there is
no clear winner between Natural exceptions and Transient exceptions
despite the theoretically advantageous additional checks of the Natural
semantics.

>>>>

Figure~\ref{fig:success-bars} summarizes the overall success rates of
every mode.\footnote{The random mode behaves the same for
all three semantics, so this analysis considers it a single mode.}
The success rates illustrate a few points that underlie the
rest of the analysis.  The first notable piece of information from this
figure is that every mode has failed debugging scenarios, not just
Erasure. This should not come as a surprise to the astute reader.  Running
a rational programmer mode on a scenario may result in an exception
that carries no useful information about which module to equip with types next. For instance the stack trace of the exception may not
contain frames from any untyped module of the program. This can happen at any point
along a blame trail, causing it to fail.

@@

Figure~\ref{fig:avo-bars} also provides answers for $Q_*$.
Blame for all three semantics is significantly
more useful than Erasure exceptions (by around 35\%). Natural blame is
more useful than both versions of Transient blame by a small percentage;
in about 4\% of scenarios Natural blame is more useful while in about 1.5\% of the
scenarios Transient blame is more useful. The Transient first and
Transient last blame are practically indistinguishable. Finally, there is
no clear winner between Natural exceptions and Transient exceptions
despite the theoretically advantageous additional checks of Natural.

### 2021-03-03 writing/why-rational-methodology.tex
touch ups in section 3
dangling lines

Every time the rational programmer debugs a program, it is validation for
programming language researchers. It shows how their slogans and theorems help
programmers connect the dots. The example of blame assignment mechanisms makes
this point clearly. A blame-assignment mechanism provides information that,
according to programming language research, points toward the source of the
problem.

@@

\item Does a language design provide information that guides the actions of the
 rational programmer?

@@

Here is how the answers to these methodological questions imagines an evaluation
method for blame-assignment mechanisms:

@@

That said, using the method to conduct data-gathering experiments entails
several challenges. The specific challenges are spelled out in the next section,
and the following three sections explain ways of overcoming them.

>>>>

Every time the rational programmer succeeds, it is validation for programming
language researchers. It shows how their theorems and slogans help
programmers>. The example of blame assignment mechanisms makes this point
clearly. A blame-assignment mechanism provides information that, according to
programming language research, points toward the source of the problem.

@@

\item Does a language design provide information that can guide a
 rational programmer?

@@

Here is how the answers to these methodological questions lead to an evaluation
method for blame-assignment mechanisms:

@@

\end{enumerate}
That said, using the method to conduct data-gathering experiments entails
several challenges. The specific challenges are spelled out in the next section,
and the following three sections explain ways of overcoming them.
